---
title: "String manipulation"
author: ""
date: ""
output: html_document
editor_options: 
  chunk_output_type: console
---

## Main ideas

- Working with string data is essential for a number of data science tasks,
  including data cleaning, data preparation, and text analysis.
  
- The `stringr` package in `R` (part of the `tidyverse`) contains useful 
  tools for working with character strings.
  

## Packages

```{r message = FALSE}
library(tidyverse)
```

## Lecture notes and exercises

`stringr` provides tools to work with character strings. Functions in `stringr`
have consistent, memorable names. 

- All begin with `str_` (`str_count()`, `str_detect()`, `str_trim()`, etc). 
- All take a vector of strings as their first arguments.
- We only have time to explore the basics. I encourage you to explore on your 
  own using the **additional resources** below.

### Preliminaries

Character strings in `R` are defined by double quotation marks. These can 
include numbers, letters, punctation, whitespace, etc.

```{r first-string}
string1 <- "CS3072 is my favorite class"
string1
```

You can combine character strings in a vector.

```{r combine-strings}
string2 <- c("CS3072", "Data Science", "Effat University")
string2
```

**Question:** What if we want to include a quotation in a string? Why doesn't
the code below work?

```{r quote-in-string, eval = FALSE}
string3 <- "I said "Hello" to my class"
```

To include a double quote in a string **escape it** using a backslash. Try it 
now in the code chunk below and name your string `string4`.

```{r quote-in-string-fixed}
string4 <- "I said \"Hello\" to my class"
```

If you want to include an actual backslash, **escape it** as shown below. This
may seem tedious but it will be important later.

```{r backslash-in-string}
string5 <- "\\"
```

The function `writeLines()` shows the content of the strings not including
escapes. Try it for `string1`, `string2`, `string3`, `string4`, and `string5` 
in the code chunk below.

```{r demo-writelines}
writeLines(string5)
```

### U.S. States

To demonstrate the basic functions from `stringr` we will use a vector of all 50
U.S. states.

```{r states-vector}
states <- c("alabama", "alaska", "arizona", "arkansas", "california", 
            "colorado", "connecticut", "delaware", "florida", "georgia", 
            "hawaii", "idaho", "illinois", "indiana", "iowa", "kansas", 
            "kentucky", "louisiana", "maine", "maryland", "massachusetts", 
            "michigan", "minnesota", "mississippi", "missouri", "montana", 
            "nebraska", "nevada", "new hampshire", "new jersey", 
            "new mexico", "new york", "north carolina", "north dakota", "ohio", 
            "oklahoma", "oregon", "pennsylvania", "rhode island",
            "south carolina", "south dakota", "tennessee", "texas", "utah", 
            "vermont", "virginia", "washington", "west virginia", "wisconsin",
            "wyoming")
```

#### `str_length()`

Given a string, return the number of characters.

```{r str-length}
string1
str_length(string1)
```

Given a vector of strings, return the number of characters in each string.

```{r str-length-vector}
str_length(states)
```

#### `str_c()`

Combine two (or more) strings.

```{r strc}
str_c("CS 3072", "is", "my", "favorite", "class")
```

Use `sep` to specify how the strings are separated.

```{r strc-sep}
str_c("CS 3072", "is", "my", "favorite", "class", sep = " ")
```

### `str_to_lower()` and `str_to_upper()`

Convert the case of a string from lower to upper or vice versa.

```{r str-cases}
str_to_upper(states)
```

### `str_sub()`

Extract parts of a string from `start` to `end`, inclusive.

```{r str-sub-1}
str_sub(states, 1, 4)
```

```{r str-sub-2}
str_sub(states, -4, -1)
```

**Practice:** Combine `str_sub()` and `str_to_upper()` to capitalize first letter of each 
state (you can ignore two word states).

```{r cap-states}
str_sub(states, 1, 1) <- str_to_upper(str_sub(states, 1, 1))
states
```

#### `str_sort()`

Sort a string. Below we sort in decreasing alphabetical order.

```{r str-sort}
str_sort(states, decreasing = TRUE)
```

### Regular Expressions

A **regular expression** is a sequence of characters that allows you to describe
string patterns. We use them to search for patterns.

Examples of usage include the following data science tasks:

- extract a phone number from text data
- determine if an email address is valid
- determine if a password has some specified number of letters, characters, 
  numbers, etc
- count the number of times "statistics" occurs in a corpus of text

To demonstrate regular expressions, we will use a vector of the states 
bordering North Carolina.

```{r nc-states}
nc_states <- c("North Carolina", "South Carolina", "Virginia", "Tennessee", 
               "Georgia")
```

#### Basic Match

We can match exactly using a **basic match**.

```{r basic-match, eval = FALSE}
str_view_all(nc_states, pattern = "in")
```

We can match any character using .

```{r any-char, eval = FALSE}
str_view_all(nc_states, pattern = ".a")
```

**Question:** What if we want to match a period .?

*Escape it* using `\.` This is the regular expression.

But we represent regular expressions using strings and \ is also an escape 
symbol in strings.

*Escape again!*

To create the regular expression `\.`, use the string "`\\.`"

```{r reg-ex-escape, eval = FALSE}
str_view(c("a.c", "abc", "def"), "a\\.c")
```

#### Anchors

Match the start of a string using ^.

```{r power-anchor, eval = FALSE}
str_view(nc_states, "^G")
```

Match the end of a string using $.

```{r money-anchor, eval = FALSE}
str_view(nc_states, "a$")
```

#### `str_detect()`

Determine if a character vector matches a pattern.

```{r str-detect}
nc_states
str_detect(nc_states, "a")
```

#### `str_subset()`

```{r str-subset}
nc_states
str_subset(nc_states, "e$")
```

#### `str_count()`

Determine how many matches there are in a string.

```{r str-count}
nc_states
str_count(nc_states, "a")
```

#### `str_replace()` and `str_replace_all()`

Replace matches with new strings.

```{r str-replace}
str_replace(nc_states, "a", "-")
```

Use `str_replace_all()` to replace all matches with new strings.

```{r str-replace-all}
str_replace_all(nc_states, "a", "-")
```

#### Many Matches

The regular expressions below match more than one character.

- Match any digit using `\d` or [[:digit:]]
- Match any whitespace using `\s` or [[:space:]]
- Match f, g, or h using [fgh]
- Match anything but f, g, or h using [^fgh]
- Match lower-case letters using [a-z] or [[:lower:]]
- Match upper-case letters using [A-Z] or [[:upper:]]
- Match alphabetic characters using [A-z] or [[:alpha:]]

Remember these are regular expressions! To match digits you'll need to escape 
the \, so use "\\d", not "\d"

## Practice

To practice manipulating strings we will use question and answer data from
two recent seasons (2008 - 2009) of the television game show *Jeopardy!*.

```{r read-in-jeopardy}
jeopardy <- read_csv("data/questions.csv")
```

- `category`: category of question
- `value`: value of question in dollars
- `question`: text of question
- `answer`: text of question answer
- `year`: year episode aired

```{r glimpse-data}
glimpse(jeopardy)
```

(1) Use a single code pipeline and a function from `stringr` to return all rows
where the answer **contains** the word "Durham"

```{r practice-1}

```

(2) Use a single code pipeline and `stringr` to find the length of all of the 
answers, sort by decreasing length, and return the five longest answers.

```{r practice-2}

```

(3) What answer has the most digits?

```{r practice-3}

```

(4) Return all rows where the category has a period.

```{r practice-4}

```

(5) Using a single code pipeline, return all rows where the question contains a 
(numeric) year between 1800 and 1999

```{r practice-5}

```

(6) Using a single code pipeline, return all rows with answers that begin with 
three vowels.

```{r practice-6}

```

(7) Using a single code pipeline, return all answers that end with `ugh` but not
`ough`.

```{r practice-7}

```

(8) Use a single code pipeline to create a new variable `prop_vowel` that
is the proportion of all letters in each answer that are vowels. What is the
highest? Lowest?

```{r practice-8}
jeopardy %>%
  mutate(vowels = str_count(answer, "[AEIOUaeiou]"),
         letters = str_count(answer, "[[:alpha:]]"),
         prop_vowel = vowels / letters) %>%
  select(answer, vowels, letters, prop_vowel) %>%
  arrange(desc(prop_vowel)) %>%
  filter(letters > 5, 
         !is.na(prop_vowel)) %>%
  slice(1:3, (n() - 2):n())
```

## Additional Resources

- [`stringr` website](https://stringr.tidyverse.org/)
- [`stringr` cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/strings.pdf)
- [Regular Expressions cheat sheet](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf)
- [R for Data Science: Strings](https://r4ds.had.co.nz/strings.html#matching-patterns-with-regular-expressions)
